---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# lmeInfo

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/jepusto/lmeInfo.svg?branch=master)](https://travis-ci.org/jepusto/lmeInfo)
[![Codecov test coverage](https://codecov.io/gh/jepusto/lmeInfo/branch/master/graph/badge.svg)](https://codecov.io/gh/jepusto/lmeInfo?branch=master)
<!-- badges: end -->

## Information matrices for fitted nlme::lme models

`lmeInfo` provides analytic derivatives and information matrices for fitted lmeStruct models from `nlme::lme()`. The package includes functions for estimating the inverse of the expected or average Fisher information matrix of the full maximum or restricted maximum likelihood. It entails estimating the sampling variance of variance parameters, including the parameters of the random effects structure, variance structure and correlation structure. Function is also available for calculating design-comparable standardized mean difference effect size (Pustejovsky et al., 2014) for multiple baseline across behaviors designs, multi-level multiple baseline designs, alternating treatment designs and meta-analyses.

## Installation

You can install the released version of lmeInfo from [CRAN](https://CRAN.R-project.org) with:

```{r eval=FALSE}
install.packages("lmeInfo")
```

You can also install the development version from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("jepusto/lmeInfo")
```

## Demonstration

We use the dataset from Bryant et al. (2016) to demonstrate how `lmeInfo` estimates the inverse Fisher information matrices for fitted LME models and calculates the design-comparable standardized mean difference effect size for a three-level multiple baseline design.

- The inverse of Fisher information matrices

```{r example}
library(lmeInfo)
library(nlme)
data(Bryant2016)

Bryant2016_RML <- lme(fixed = outcome ~ treatment,
                      random = ~ 1 | school/case,
                      correlation = corAR1(0, ~ session | school/case),
                      data = Bryant2016)
```

The inverse of the expected Fisher information matrix of model `Bryant2016_RML` can be estimated using `Fisher_info()` in `lmeInfo` when the type argument is specified as __expected__. The inverse of the average Fisher information matrix is estimated when type is specified as __average__.

```{r, eval = FALSE}
Fisher_info(Bryant2016_RML, type = "expected")
#>                             Tau.school.var((Intercept)) Tau.case.var((Intercept))    cor_params      sigma_sq
#> Tau.school.var((Intercept))                1.322700e-05              3.481279e-06  4.031958e-02  2.034762e-06
#> Tau.case.var((Intercept))                  3.481279e-06              1.299374e-05  1.510332e-01  7.536162e-06
#> cor_params                                 4.031958e-02              1.510332e-01  2.744559e+05 -5.762687e+00
#> sigma_sq                                   2.034762e-06              7.536162e-06 -5.762687e+00  1.320509e-04
```

- The design-comparable standardized mean difference effect size

```{r}
Bryant2016_g <- g_mlm(Bryant2016_RML, p_const = c(0,1), r_const = c(1,1,0,1))
```

```{r}
print(Bryant2016_g)
```

Pustejovsky, Hedges & Shadish (2014) provides detailed estimation methods for the effect sizes, standard errors and degree of freedom.

## References

Bryant, B. R., Bryant, D. P., Porterfield, J., Dennis, M. S., Falcomata, T., Valentine, C., ... & Bell, K. (2016). The effects of a Tier 3 intervention on the mathematics performance of second grade students with severe mathematics difficulties. Journal of learning disabilities, 49(2), 176-188. doi:10.1177/0022219414538516

Pustejovsky, J. E., Hedges, L. V., & Shadish, W. R. (2014). Design-comparable effect sizes
in multiple baseline designs: A general modeling framework. Journal of Educational
and Behavioral Statistics, 39 (5), 368–393. doi:10.3102/1076998614547577
