---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# lmeInfo

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/jepusto/lmeInfo.svg?branch=master)](https://travis-ci.org/jepusto/lmeInfo)
[![Codecov test coverage](https://codecov.io/gh/jepusto/lmeInfo/branch/master/graph/badge.svg)](https://codecov.io/gh/jepusto/lmeInfo?branch=master)
<!-- badges: end -->

## Information matrices for fitted nlme::lme models

`lmeInfo` provides analytic derivatives and information matrices for fitted lmeStruct models from `nlme::lme()`. The package includes functions for estimating the sampling variance-covariance of variance component parameters using the inverse Fisher information for fitted LME models. The variance component paramters include the parameters of the random effects structure, the variance structure and the correlation structure. The package entails the calculation of the inverse of the expected or average Fisher information matrix of the full maximum or restricted maximum likelihood. Function is also available for calculating design-comparable standardized mean difference effect size (Pustejovsky et al., 2014) for multiple baseline across behaviors designs, multi-level multiple baseline designs, alternating treatment designs and meta-analyses.

## Installation

You can install the released version of lmeInfo from [CRAN](https://CRAN.R-project.org) with:

```{r eval=FALSE}
install.packages("lmeInfo")
```

You can also install the development version from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("jepusto/lmeInfo")
```

## Demonstration

We use the dataset from Bryant et al. (2016) to demonstrate how `lmeInfo` estimates the sampling variance-covariance of variance component parameters of fitted LME models and calculates the design-comparable standardized mean difference effect size for a three-level multiple baseline design.

### Sampling variance-covariance of variance component parameters

```{r example}
library(lmeInfo)
library(nlme)
data(Bryant2016)

Bryant2016_RML <- lme(fixed = outcome ~ treatment,
                      random = ~ 1 | school/case,
                      correlation = corAR1(0, ~ session | school/case),
                      data = Bryant2016)
```

The sampling variance-covariance of variance component parameters of model `Bryant2016_RML` can be estimated using `varcomp_vcov()` in `lmeInfo`. The sampling variance-covariance of variance component parameters is calculated using the inverse expected Fisher information when the type argument in `varcomp_vcov()` is specified as __expected__. The inverse of the average Fisher information matrix is estimated when the type argument is specified as __average__.

```{r, eval = FALSE}
varcomp_vcov(Bryant2016_RML, type = "expected")
#>                             Tau.school.var((Intercept)) Tau.case.var((Intercept))    cor_params      sigma_sq
#> Tau.school.var((Intercept))                8.133900e+04             -2.140289e+04 -0.0100434219 -4.701744e+02
#> Tau.case.var((Intercept))                 -2.140289e+04              3.996251e+05 -8.2280433012 -3.815479e+05
#> cor_params                                -1.004342e-02             -8.228043e+00  0.0002154736  9.872975e+00
#> sigma_sq                                  -4.701744e+02             -3.815479e+05  9.8729751627  4.602107e+05
```

### The design-comparable standardized mean difference effect size

The design-comparable standardized mean difference effect size is calculated using `g_mlm()` in `lmeInfo`. The standard errors of the unadjusted or adjusted effect size are estimated with the inverse expected Fisher information matrix when the infotype argument in `g_mlm()` is specified as __expected__. The inverse average Fisher information matrix is used when the infotype argument is specified as __average__. The estimation method is built on the method described in Pustejovsky, Hedges, & Shadish (2014).

```{r}
Bryant2016_g <- g_mlm(Bryant2016_RML, p_const = c(0,1), r_const = c(1,1,0,1), infotype = "expected")
```

```{r}
print(Bryant2016_g)
```

## References

Bryant, B. R., Bryant, D. P., Porterfield, J., Dennis, M. S., Falcomata, T., Valentine, C., ... & Bell, K. (2016). The effects of a Tier 3 intervention on the mathematics performance of second grade students with severe mathematics difficulties. Journal of learning disabilities, 49(2), 176-188. doi:10.1177/0022219414538516

Pustejovsky, J. E., Hedges, L. V., & Shadish, W. R. (2014). Design-comparable effect sizes
in multiple baseline designs: A general modeling framework. Journal of Educational
and Behavioral Statistics, 39 (5), 368–393. doi:10.3102/1076998614547577
